Seeing Double: Indications From {\it TESS} and {\it Gaia} That PTFO$\,$8-8695b is Not a Planet

### 2020-04-16 paper/ms.tex
paper smithing

Additional lines of evidence could imply further fascinating ``firsts'' for
this candidate planet.  One first would be that its transits are about three
times deeper in optical bandpasses ({\it e.g.,} $g$) than in the near-infrared
({\it e.g.}, $z$-band) \citep{onitsuka_multicolor_2017,tanimoto_evidence_2020}.
A cloud-free hydrogen-dominated planetary atmosphere cannot explain such a
wavelength dependence.  The proposed explanation is that the planet is
surrounded by a dust cloud \citep{tanimoto_evidence_2020}.

A second first could be the detection of direct H$\alpha$ emission from the
planet itself \citep{johnskrull_h_2016}.  While the stellar chromosphere emits
in H$\alpha$, it seems that there is an additional excess H$\alpha$ emission
that could be in phase with the planetary orbit.  The average velocity width of
the excess H$\alpha$ emission is 87$\,$km$\,$s$^{-1}$, and its equivalent width
is 70-80\% that from the stellar chromosphere \citep{johnskrull_h_2016}.  The
proposed explanation is that the outflowing mass from the planet may explain
this excess emission as well \citep{johnskrull_h_2016}.

There are perhaps a few challenges to the planetary interpretation.  They
include that the planet does not seem to emit infrared radiation in occulation,
at least anywhere near the expected amplitude \citep{yu_tests_2015}.  In
addition, despite measurement attempts by multiple investigators, it does not
seem to show the Rossiter effect at the amplitude expected given the rapid
stellar rotation and large planet size
\citep{yu_tests_2015,ciardi_followup_2015}.  Finally, detailed modelling of the
``precession + gravity darkening'' transits has shown that the necessary degree
of gravity darkening is too great, given the spectroscopically observed
equatorial velocity \citep{howarth_reappraisal_2016}.  Additionally, the
gravity-darkened star would need to show photometric variability from its
precession rotation axis, and such variability has not been observed.

While the planetary interpetation has challenges, so too do alternative
explanations.  High-latitude accretion hotspots might produce the observed
H$\alpha$ variability, but \ptfo\ does not have an infrared (IR) excess
associated with the presence of an inner disk.  Low-latitude spots, hot or
cold, struggle to produce the necessary dip durations.  A circumstellar cloud
of dust or gas might be viable, but clumps must be invoked to explain the
short-duration flux dips.  The material might be a remnant of the primordial
gas disk, or it could be ejected and entrained in the stellar magnetosphere.

Section~\ref{sec:observations} presents newly available observations from the
Transiting Exoplanet Survey Satellite (TESS; \citealt{ricker_transiting_2015})
and Gaia \citep{gaia_collaboration_gaia_2018}.  Analyzing the TESS data in
Section~\ref{sec:tess}, we find two distinct signals.  A long sinusoid repeats
every 12.0 hours, and is probably stellar rotation.  A short ``dip + complex
modulation'' repeats every 10.7 hours.  Analyzing the Gaia data in
Section~\ref{sec:gaia}, we show that \ptfo\ is quite likely a photometric
binary.  In Section~\ref{sec:discussion}, we note that the orbital phase of the
dip seems to have instantaneously changed, perhaps twice.  We argue that
PTFO$\,$8-8695 is a binary M dwarf in which one star shows a rotation signal,
and the other is showing ``transient dipping'' that has been observed in other
young M dwarfs, and is likely caused by eclipses of warm particulate clouds.
Section~\ref{sec:conclusions} summarizes our main points.

@@

In describing the cluster membership of \ptfo, we follow the notation
and results of \citet{kounkel_apogee2_2018}.
\citet{kounkel_apogee2_2018} combined astrometric data from Gaia DR2
with spectroscopic data from APOGEE-2 (CITE).  They then performed a
hierarchical clustering on the six dimensional position and velocity
information to identify subgroups within the Orion complex.  From
smallest to largest groups, \ptfo\ was identified as being a member of
the following subgroups:
\begin{equation}
  {\rm PTFO\,8\text{-}8695}
  \in {\rm 25\,Ori\text{-}1}
  \subset {\rm 25\,Ori}
  \subset {\rm Orion\ OB1a}
  \subset {\rm Orion\ D},
\end{equation}
where from set-notation, `$\subset$' denotes ``is a proper subset of''.

@@

% \begin{figure*}[t]
% 	\begin{center}
% 		\leavevmode
% 		\includegraphics[width=0.9\textwidth]{f6.png}
% 	\end{center}
% 	\vspace{-0.7cm}
% 	\caption{ {\bf Positions, kinematics, and HR diagram of \ptfo.}
%   Members of the 25$\,$Ori-1 subgroup are shown with black circles,
%   and were identified by \citet{kounkel_apogee2_2018} through
%   clustering on Gaia-DR2 and APOGEE data.  The ``neighborhood'' (gray
%   circles) is defined as the group of at most $10^4$ randomly selected
%   non-member stars within 5 standard deviations of the mean right
%   ascension, declination, and parallax.
% 	\label{fig:gaia}
% 	}
% \end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f6.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf HR diagram of \ptfo\ and members of 25$\,$Ori-1.}
  Members of the 25$\,$Ori-1 group (black circles) were identified by
  \citet{kounkel_apogee2_2018} through clustering on six-dimensional
  Gaia-DR2 and APOGEE-2 data.  We defined the reference
  ``neighborhood'' (gray circles) as the group of at most $10^4$
  randomly selected non-member stars within 5 standard deviations of
  the mean 25$\,$Ori-1 right ascension, declination, and parallax.
  \ptfo\ seems likely to be a photometric binary.
  $G$ is the Gaia broadband, $Bp$ is Gaia blue, $Rp$ is Gaia red, and
  $\omega_{\rm as}$ is the parallax in arcseconds.
	\label{fig:gaia}
	}
\end{figure*}

>>>>

Other lines of evidence would imply further planetary ``firsts'' for
this planet candidate.  One first would be that its transits are about three
times deeper in optical bandpasses ({\it e.g.,} $g$-band) than in the near-infrared
({\it e.g.}, $z$-band) \citep{onitsuka_multicolor_2017,tanimoto_evidence_2020}.
A cloud-free hydrogen-dominated planetary atmosphere cannot explain such a
wavelength dependence.  The planet might therefore be surrounded by a
dust cloud \citep{tanimoto_evidence_2020}.

Another first could be the direct detection of H$\alpha$ emission from the
planet itself \citep{johnskrull_h_2016}.  While the stellar chromosphere emits
in H$\alpha$, it seems that there is an additional excess H$\alpha$ emission
that could be in phase with the planetary orbit.  The average velocity width of
the excess H$\alpha$ emission is 87$\,$km$\,$s$^{-1}$, and its equivalent width
is 70-80\% that of the stellar chromosphere \citep{johnskrull_h_2016}.  The
proposed explanation is that the outflowing mass from the planet may explain
this excess emission as well \citep{johnskrull_h_2016}.

There are perhaps a few challenges to the planetary interpretation (if
these ``features'' are not already seen as such).  They include that
the planet does not seem to emit infrared radiation in occulation, at
least anywhere near the expected amplitude \citep{yu_tests_2015}.  In
addition, despite measurement attempts by multiple investigators, it
does not seem to show the Rossiter effect at the amplitude expected
given the rapid stellar rotation and large planet size
\citep{yu_tests_2015,ciardi_followup_2015}.  Finally, detailed
modelling of the ``precession + gravity darkening'' transits has shown
that the necessary degree of gravity darkening is too great, given the
spectroscopically observed equatorial velocity
\citep{howarth_reappraisal_2016}.  Additionally, as the gravity-darkened
star precessed about its rotation axis, it would need to show
photometric variability that has not been observed.

While the planetary interpetation has challenges, alternative
explanations do as well.  High-latitude accretion hotspots might
produce the observed H$\alpha$ variability, but \ptfo\ does not have
an infrared (IR) excess associated with the presence of an inner disk.
Low-latitude starspots, hot or cold, struggle to produce the necessary
dip durations.  A circumstellar cloud of dust or gas might be
entrained in the stellar magnetosphere, but clumps would need to be
invoked to explain the short-duration flux dips.

In the following, Section~\ref{sec:observations} presents newly
available observations from the Transiting Exoplanet Survey Satellite
(TESS; \citealt{ricker_transiting_2015}) and Gaia
\citep{gaia_collaboration_gaia_2018}.  Analyzing the TESS data in
Section~\ref{sec:tess}, we find two distinct signals.  A long sinusoid
repeats every 12.0 hours, and is probably stellar rotation.  A short
``dip + complex modulation'' repeats every 10.7 hours.  Analyzing the
Gaia data in Section~\ref{sec:gaia}, we show that \ptfo\ is quite
likely a photometric binary.  In Section~\ref{sec:discussion}, we note
that the orbital phase of the dip seems to have instantaneously
changed, perhaps twice.  We argue that PTFO$\,$8-8695 is a binary M
dwarf in which one star shows a rotation signal, and the other is
showing ``transient dipping'' that has been observed in other young M
dwarfs, and is likely caused by eclipses of warm particulate clouds.
Section~\ref{sec:conclusions} summarizes our main points.

@@

In describing the cluster membership of \ptfo, we follow the notation
and results of \citet{kounkel_apogee2_2018}.
\citet{kounkel_apogee2_2018} combined astrometric data from Gaia DR2
with spectroscopic data from APOGEE-2 (CITE).  They then performed a
hierarchical clustering on the six dimensional position and velocity
information to identify subgroups within the Orion complex.  From
smallest to largest groups, \ptfo\ was identified as being a member of
the following nested subgroups:
\begin{equation}
  %{\rm PTFO\,8\text{-}8695}
  % \in
  {\rm 25\,Ori\text{-}1}
  \subset {\rm 25\,Ori}
  \subset {\rm Orion\ OB1a}
  \subset {\rm Orion\ D},
\end{equation}
where from set-notation, `$\subset$' denotes ``is a proper subset of''.

@@

% \begin{figure*}[t]
% 	\begin{center}
% 		\leavevmode
% 		\includegraphics[width=0.9\textwidth]{f6.png}
% 	\end{center}
% 	\vspace{-0.7cm}
% 	\caption{ {\bf Positions, kinematics, and HR diagram of \ptfo.}
%   Members of the 25$\,$Ori-1 subgroup are shown with black circles,
%   and were identified by \citet{kounkel_apogee2_2018} through
%   clustering on Gaia-DR2 and APOGEE data.  The ``neighborhood'' (gray
%   circles) is defined as the group of at most $10^4$ randomly selected
%   non-member stars within 5 standard deviations of the mean right
%   ascension, declination, and parallax.
% 	\label{fig:gaia}
% 	}
% \end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f6.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf HR diagram of \ptfo\ and late-type members of 25$\,$Ori-1.}
  Members of the 25$\,$Ori-1 group (black circles) were identified by
  \citet{kounkel_apogee2_2018} through clustering on six-dimensional
  Gaia-DR2 and APOGEE-2 data.  We defined the reference
  ``neighborhood'' (gray circles) as the group of at most $10^4$
  randomly selected non-member stars within 5 standard deviations of
  the mean 25$\,$Ori-1 right ascension, declination, and parallax.
  $G$ is the Gaia broadband, $Bp$ is Gaia blue, $Rp$ is Gaia red, and
  $\omega_{\rm as}$ is the parallax in arcseconds.
  The $x$-axis limits have been set to show only K and M dwarf
  members, to accentuate \ptfo's separation from the
  single-star sequence.
	\label{fig:gaia}
	}
\end{figure*}
